Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-22 paper/ms.tex
edits to conclusion

\newcommand{\foreign}[1]{\emph{#1}}
\newcommand{\etal}{\foreign{et\,al.}}
\newcommand{\etc}{\foreign{etc.}}
\newcommand{\ie}{\foreign{i.e.}}

@@

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are a class of stochastic
models consisting of a mean function $\mu_\bvec{\theta}(\bvec{x})$ and a
covariance, autocorrelation, or ``kernel'' function
$k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ parameterized by the parameters
$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, the log-likelihood of observing a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.

GP models have been widely used across the physical sciences but their
application is generally limited to small datasets because the cost of
computing the inverse and determinant of the matrix $K_\bvec{\alpha}$ is
$\mathcal{O}(N^3)$.
In other words, this cost is proportional to the cube of the number of data
points $N$.
This means that for large datasets, every evaluation of the likelihood
quickly becomes computationally intractable.
In this case, the use of non-linear optimization or MCMC is no longer
practical.

In the following Section, we present a method for substantially improving this
scaling in many circumstances.
We call our method and its implementations \celerite.\footnote{The name
\celerite\ comes from the French word \foreign{c\'elerit\'e} meaning the speed
of light in a vacuum.} The \celerite\ method requires using a specific model
for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and, although it
has some limitations, we demonstrate in subsequent sections that it can
increase the computational efficiency of many astronomical data analysis
problems.
The main limitation of this method is that it can only be applied to
one-dimensional datasets, where by ``one-dimensional'' we mean that the
 \emph{input coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv
t_n$.\footnote{We are using $t$ as the input coordinate because
one-dimensional GPs are often applied to time series data but this isn't a
real restriction and the \celerite\ method can be applied to \emph{any}
one-dimensional dataset.}
Furthermore, the covariance function for the \celerite\ method is
``stationary'', \ie\ $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function of
$\tau_{nm} \equiv |t_n - t_m|$.

@@

The computational burden of Gaussian process regression has limited its application
to small astronomical time series.  With the linear scaling of \celerite\, we
envision the application of Gaussian processes may be expanded to much larger
datasets.  Despite the restrictive form of the \celerite\ kernel, with a
sufficient number of components it is flexible enough to describe a wide
range of astrophysical variability.  In fact, the relation of the \celerite\ kernel
to the damped, driven harmonic oscillator matches simple models of astrophysical
variability, and makes the parameterization interpretable in terms of resonant
frequency, amplitude, and quality factor.  Perhaps the most significant drawback
is the $J^2$ scaling of the solver;  however, for many problems a small value of $J$
suffices.

We envision two general varieties of applications:  those interested in the
the \celerite\ kernel parameters (so-called ``hyper-parameters") which describe the
astrophysical variability, versus those interested in characterizing the mean
model, treating the correlated noise as a nuisance.  Some users may
be interested in both applications, while other problems may require interpolation
or extrapolation beyond the dataset that is being modeled.

%Some examples of the first are determining the characteristic variability
%timescale of a quasar \citep{Kelly:2014}, measuring the asteroseismic
%variation of a star \citep{Brewer:2009,Corsaro:2014}, measurement of
%granulation noise \citep{Bastien:2013,Kallinger:2016}, detection of
%quasi-periodic variability in a high-energy source \citep{McAllister:2016}, or
%classification of variable objects \citep{Zinn:2016}.
%Examples of the second are detecting and characterizing transiting exoplanets
%\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016,
%Grunblatt:2016, Luger:2016}
%and correcting for stellar activity in radial velocity measurements
%\citep[for example,][]{Haywood:2014, Rajpaul:2015}.
%Examples of the third are measuring time delays in multiply-imaged
%gravitationally lensed sources \citep{Tewes:2013} and reverberation mapping of
%AGN.

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt full covariance matrices in analyzing the noise in
stellar lightcurves when detecting or characterizing transiting planets
\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016,
Grunblatt:2016, Luger:2016}.
One promising early attempt at accounting for correlated noise was the wavelet
approach of \citet{Carter:2009}.
Their technique allowed for a power-law spectrum for the noise scaling as
$\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
physically-accurate, description of stellar variability.
Their approach runs in ${\cal O}(N)$ time for $\alpha=1$, which is referred to
as `flicker', $1/f$, or pink noise;  however, this particular type of noise
overpredicts the power spectrum at low and high frequency compared to the Solar
granulation power spectrum.  It also cannot describe quasi-periodic noise.

Further progress was made by parameterizing the covariance matrix with simple,
analytic functions that describe the autocorrelation function \citep{Gibson:2012}.
The parameters of these functions can then be optimized to best match the observed
variability pattern of the residuals (after subtracting a transit model).  The
disadvantage of this approach is that the functions used are chosen somewhat
arbitrarily (e.g.\ the exponential-squared function), and the Cholesky decomposition
of the covariance matrix for computing the likelihood takes ${\cal O}(N^3)$ operations
which prohibits application of this technique to large datasets.

Given the drawbacks of these approaches, the \celerite\ formalism allows both a fast
computation of the likelihood in ${\cal O}(N)$ time, as well as a functional form
that accurately describes stellar variability due to the relation to the simple
harmonic oscillator power spectrum, which is an analog of stellar asteroseismic
oscillations.  As higher signal-to-noise observations of transiting exoplanet systems
are obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters, and so we expect that \celerite\
will be important for transit detection \citep{ForemanMackey:2015}, transit timing
\citep{Agol:2005,Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler
beaming \citep{Loeb:2003,Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase
functions \citep{Knutson:2007,Zucker:2007}, and more.

Although we have in mind applications of this fast method to model stellar
variability, the method is general for one-dimensional GP problems, and may be
applied to other problems.
Accreting black holes show time series which may be modeled by correlated
noise \citep{Kelly:2014};  indeed, this was the motivation for the original
technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
This approach may be broadly used for characterizing quasar variability
\citep{MacLeod:2010}, measuring time lags with reverberation mapping
\citep{Zu:2011,Pancoast:2014}, modeling time delays in multiply-imaged
gravitationally-lensed systems \citep{Press:1998}, characterizing quasi-periodic
variability in a high-energy source \citep{McAllister:2016}, or
classification of variable objects \citep{Zinn:2016}.
Outside of astronomy, this technique may have application to terrestrial
seismology \citep{Robinson:1967}.
%The \celerite\ kernel may be used in lieu of a Gaussian power-spectrum for
%non-parametric data modeling \citep{Wilson:2013}.

The \celerite\ formalism also allows a principled approach to the estimation
of the power spectrum of a time series, as well as its uncertainties.
Fourier transforms work for uniformly sampled data;  astronomical data are
rarely (if ever) uniformly sampled.  Periodograms are affected by the window
function and aliasing in datasets, and make estimating the significance of
spectral features ambiguous.  The Bayesian approach to the estimation of a power
spectrum \citep{Wilson:2013} avoids these difficulties:  it allows for the computation of
uncertainties, it accounts for the window function on the uncertainties of
the power spectrum, and in principle it can treat aliases if the full posterior
is explored.  One drawback is that the number of $J$ terms to include is
unknown, but may be inferred with computation of the evidence, or an
approximation such as the Bayesian information criterion \citep{Kelly:2014}.
In general, the parameters of the \celerite\ terms are more interesting due
to their physical meaning rather than the overall shape of the power spectrum,
as shown in the foregoing examples.

>>>>

\newcommand{\foreign}[1]{\emph{#1}}
\newcommand{\etal}{\foreign{et\,al.}}
\newcommand{\etc}{\foreign{etc.}}

@@

GP are stochastic models consisting of a mean function
$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ parameterized by the
parameters $\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, the log-likelihood of observing a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.

GP models have been widely used across the physical sciences but their
application is generally limited to small datasets because the computational
cost of computing the inverse and determinant of the matrix $K_\bvec{\alpha}$
is $\mathcal{O}(N^3)$.
In other words, this cost is proportional to the cube of the number of data
points $N$.
This means that for large datasets, every evaluation of the likelihood
quickly becomes computationally intractable.
In this case, the use of non-linear optimization or MCMC is no longer
practical.

In the following Section, we present a method of substantially improving this
scaling in many circumstances.
We call our method and its implementations \celerite.\footnote{The name
\celerite\ comes from the French word \foreign{c\'elerit\'e} meaning the speed
of light in a vacuum.} The \celerite\ method requires using a specific model
for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and it has
several limitations but, in subsequent sections, we demonstrate that it can be
used to increase the computational efficiency of many astronomical data
analysis problems.
The main limitation of this method is that it can only be applied to
one-dimensional datasets.
When we say ``one-dimensional'' here, it means that the \emph{input
coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv t_n$.\footnote{We are
using $t$ as the input coordinate because one-dimensional GPs are often
applied to time series data but this isn't a real restriction and the \celerite\
method can be applied to \emph{any} one-dimensional dataset.}
Furthermore, the covariance function for the \celerite\ method is ``stationary''.
This means that the function $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function
of $\tau_{nm} \equiv |t_n - t_m|$.

@@

Gaussian Process models have been fruitfully applied to many problems in
astronomical data analysis but the fact that the computational cost scales as
the cube of the number of data points has limited their use to relatively
small datasets.
With the linear scaling of \celerite\, we envision the application of Gaussian
processes may be expanded to much larger datasets.
Despite the restrictive form of the \celerite\ kernel, with a sufficient
number of components it is flexible enough to describe a wide range of
astrophysical variability.
In fact, the relation of the \celerite\ kernel to the damped, driven harmonic
oscillator matches simple models of astrophysical variability, and makes the
parameterization interpretable in terms of resonant frequency, amplitude, and
quality factor.
A major drawback of this method is its quadratic scaling with the number of
terms $J$ but, in many cases, small values of $J$ are sufficient.

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt full covariance matrices in analyzing the noise in
stellar light curves when detecting or characterizing transiting planets
\citep[for example,][]{Carter:2009, Gibson:2012, Barclay:2015, Evans:2015,
Aigrain:2016, Foreman-Mackey:2016b, Grunblatt:2016, Luger:2016}.
All of these analyses have been limited to small datasets or restrictive
kernel choices.
\celerite\ weakens these requirements by providing a scalable method for
computing the likelihood and a physical motivation for the choice of kernel.
\celerite\ can be used to accurately model stellar oscillations using the
relation to the mixture of stochastically driven, damped simple harmonic
oscillators.
As higher signal-to-noise observations of transiting exoplanet systems are
obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters.
We expect that \celerite\ will be important for transit detection
\citep{Pope:2016, Foreman-Mackey:2016b}, transit timing \citep{Agol:2005,
Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler beaming
\citep{Loeb:2003, Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase
functions \citep{Knutson:2007, Zucker:2007}, and more.

Beyond these applications to model stellar variability, the method is
generally applicable to other one-dimensional GP models.
Accreting black holes show time series which may be modeled using a GP
\citep{Kelly:2014}; indeed, this was the motivation for the original technique
developed by Rybicki \& Press \citep{Rybicki:1992, Rybicki:1995}.
This approach may be broadly used for characterizing quasar variability
\citep{MacLeod:2010}, measuring time lags with reverberation mapping
\citep{Zu:2011, Pancoast:2014}, modeling time delays in multiply-imaged
gravitationally-lensed systems \citep{Press:1998}, characterizing
quasi-periodic variability in a high-energy source \citep{McAllister:2016}, or
classification of variable objects \citep{Zinn:2016}.
We expect that there are also applications beyond astronomy.

The \celerite\ formalism can also be used for power spectrum estimation and
quantification of its uncertainties.
In principle, a large number of \celerite\ terms could be used to perform
non-parametric probabilistic inference of the power spectrum despite
unevenly-spaced data with heteroscedastic noise \citep[for
example,][]{Wilson:2013, Kelly:2014}.
This type of analysis will be limited by the quadratic scaling of \celerite\
with the number of terms $J$ but this limits existing methods as well
\citep[CARMA models,][]{Kelly:2014}.
In general, we encourage the use of physically motivated models for parameter
estimation instead of qualitative modeling of the power spectrum itself.

There are many data analysis problems where \celerite\ will not be applicable.
In particular, any multidimensional
